A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-08-03 paper/paper.tex
started revision of the paper.

The role of critical vertices is central in our algorithm
%\footnote{If the focus is not on the critical vertices for a large graph, then an algorithm that is based on randomly selecting a vertex to move, so to improve the crossing resolution, will need a large number of iterations to converge to a good solution, because it is very unlike to select one of them to move.}
\footnote{If the focus is not on the critical vertices for a large graph, then our algorithm will need a large number of iterations to converge to a good solution, because it is simply very unlikely to select to move one of the vertices that define the crossing resolution.}: By appropriately changing the location of a critical vertex or of a vertex in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a weighted random selection procedure, so that the vertices at distance~$i$ from the ones of the vertex-pool have higher weights than the corresponding ones at distance $j$  in the graph, when $0 \leq i<j$. So, if the vertex-pool contains critical vertices, then the closer a vertex is to the critical vertices, the more likely it is to be chosen. Otherwise, each vertex can be chosen with the same probability.

@@

So far, we have described the main idea of our algorithm, which at each iteration chooses uniformly at random a vertex of the current drawing to move (based on the content of the vertex-pool), so to improve the crossing resolution. Next, we  described how to compute its new position in the next drawing.

@@

% ==================================================================
\myparagraph{Avoiding local maxima}
% ==================================================================
%
To avoid getting trapped to locally optimal solutions, we mainly investigated two approaches, which are both  parametrizable by two input parameters $\zeta$ and $\zeta'$. The first mimics the human behaviour. What would~one do to escape from a locally optimal solution? She would stop trying to move the endvertices of the edges defining the crossing resolution; she would rather start moving ``irrelevant'' vertices hoping that
%by changing the embedding of the graph or its crossing structure,
by doing so a better solution will be easier to be computed afterwards. Our algorithm is mimicking this idea as follows:
%
\begin{inparaenum}[(i)]
%\item As long as the algorithm detects that within the last $\zeta$ iterations the crossing resolution of the graph has being improved, the vertex-pool contains only critical vertices.
\item if during the last $\zeta$ iterations the crossing resolution has not been~improved, then the vertex-pool becomes \emph{wider} containing all the vertices, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations;
\item afterwards, the vertex-pool switches back to the critical vertices.
\end{inparaenum}
%
%From our extensive practical analysis, we quickly realized that, while
While this approach turned out to be quite effective for medium-size graphs, for larger graphs, unfortunately, it was not so efficient; in most iterations with the wider vertex-pool, the embedding could not change in a beneficial way for the algorithm~to~proceed.

@@

% ==================================================================
\myparagraph{Complexity issues}
% ==================================================================
%
A factor that highly affects the efficiency of our algorithm is the computation of the crossing points of the edges and the corresponding angles at these points. Given  a drawing, a naive approach to compute its crossings requires $O(m^2)$ time, which can be improved by a plane-sweep technique to $O(m \log m + c)$ time, where $m$ and $c$ denote the number of edges and crossings.
%; see, e.g.,~\cite{DBLP:books/lib/BergCKO08}.

@@

% ==================================================================
\myparagraph{Total resolution}
% ==================================================================
%
%The notion of the total resolution of a drawing was introduced relatively recently with aim of ``balancing'' the measures of the crossing and of the angular resolution of a drawing~\cite{DBLP:journals/cj/ArgyriouBS13}. Formally, it is defined as the minimum of these two measures. It was not difficult to
Similarl as above, we could adjust our algorithm to yield drawings with high total resolution by simply taking into account also the angular resolution of the drawing. In particular, if the total resolution of the drawing is defined by its angular resolution, then the way we compute the critical vertices of this drawing has to change; the critical vertices must be the endvertices of the pairs of edges that define the angular resolution. Also, at each iteration of our algorithm we have to ensure that the total resolution does not decrease. We do so by rejecting candidate positions which yield a reduced total~resolution.

@@

In this section, we present the results of our experimental evaluation. For comparison purposes, apart from our algorithm, we also implemented the force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. The implementations were in Java using yFiles~\cite{DBLP:books/sp/04/WieseE004}. The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM.
%
As a test set for our experiment, we used the non-planar Rome graphs~\cite{DBLP:reference/crc/BattistaD13}, which form a collection of around 8.100 benchmark graphs; in Appendix~\ref{app:north}, we also report on the AT\&T graphs.

@@

% ==================================================================
\myparagraph{Discussion}
%
% ==================================================================
%
%In this paper, we introduced a new heuristic aiming to produce drawings of high crossing resolution, for which we also presented variants that take into account other common aesthetic criteria in Graph Drawing. Our experimental evaluation indicates that the new heuristic is competitive to state of the art force-directed algorithms, even when restricted to a given maximum aspect ratio. As a future direction, we plan to evaluate the performance of variants of our heuristic that compromise between even more aesthetic criteria.
%
While working on this project, we made some useful observations and  obtained some interesting insights. In particular, there is a recent hypothesis (also supported by experiments) that drawings, in which the crossing angles, are large are easy to read and understand. What we observed is that the drawings that are optimized only in terms of the crossing angles might be arbitrarily bad and may have several undesired properties. In particular, in these drawings it was very common to have adjacent edges to run almost in parallel and vertices to be very close to each other (nearly at the same spot). Hence, the angular resolution and the aspect ratio were very often poor. The additional restrictions that we imposed regarding the angular resolution and the aspect ratio helped in significantly improving the readability of the drawings, without loosing too much of their quality in terms of the crossing resolution.

>>>>

The role of critical vertices is central in our algorithm
%\footnote{If the focus is not on the critical vertices for a large graph, then an algorithm that is based on randomly selecting a vertex to move, so to improve the crossing resolution, will need a large number of iterations to converge to a good solution, because it is very unlike to select one of them to move.}
\footnote{If the focus is not on the critical vertices for a large graph, then our algorithm will need a large number of iterations to converge to a good solution, because it is simply very unlikely to select to move one of the vertices that define the crossing resolution.}: By appropriately changing the location of a critical vertex or of a vertex in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a weighted random selection procedure, so that the vertices at distance~$i$ from the ones of the vertex-pool (that is the shortest path to any vertex in the vertex-pool has length~$i$)  have higher weights than the corresponding ones at distance $j$  in the graph, when $0 \leq i<j$. So, if the vertex-pool contains only critical vertices, then the closer a vertex is to the critical vertices, the more likely it is to be chosen. Otherwise, the vertex-pool contains all vertices and each vertex can be chosen with the same probability.

@@

So far, we have described the main idea of our algorithm, which at each iteration chooses uniformly at random a vertex of the current drawing to move (based on the content of the vertex-pool), so to improve the crossing resolution. Next, we  describe how to compute its new position in the next drawing.

@@

% ==================================================================
\myparagraph{Avoiding local maxima}
% ==================================================================
%
To avoid getting trapped to locally optimal solutions, we mainly investigated two approaches, which are both  parametrizable by two input parameters $\zeta$ and $\zeta'$. The first mimics the human behaviour. What would~one do to escape from a locally optimal solution? She would stop trying to move the endvertices of the edges defining the crossing resolution; she would rather start moving ``irrelevant'' vertices hoping that
%by changing the embedding of the graph or its crossing structure,
by doing so a better solution will be easier to be computed afterwards. Our algorithm is mimicking this idea as follows:
%
\begin{inparaenum}[(i)]
%\item As long as the algorithm detects that within the last $\zeta$ iterations the crossing resolution of the graph has being improved, the vertex-pool contains only critical vertices.
\item if during the last $\zeta$ iterations the crossing resolution has not been~improved, then the vertex-pool becomes \emph{wider} by including all the vertices, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations;
\item afterwards, the vertex-pool switches back to the critical vertices.
\end{inparaenum}
%
%From our extensive practical analysis, we quickly realized that, while
While this approach turned out to be quite effective for medium-size graphs, for larger graphs, unfortunately, it was not so efficient; in most iterations with the wider vertex-pool, the embedding could not change in a beneficial way for the algorithm~to~proceed.

@@

% ==================================================================
\myparagraph{Complexity issues}
% ==================================================================
%
A factor that highly affects the efficiency of our algorithm is the computation of the crossing points of the edges and the corresponding angles at these points. Given  a drawing, a na\"ive approach to compute its crossings requires $O(m^2)$ time, which can be improved by a plane-sweep technique to $O(m \log m + c)$ time, where $m$ and $c$ denote the number of edges and crossings.
%; see, e.g.,~\cite{DBLP:books/lib/BergCKO08}.

@@

% ==================================================================
\myparagraph{Total resolution}
% ==================================================================
%
%The notion of the total resolution of a drawing was introduced relatively recently with aim of ``balancing'' the measures of the crossing and of the angular resolution of a drawing~\cite{DBLP:journals/cj/ArgyriouBS13}. Formally, it is defined as the minimum of these two measures. It was not difficult to
Similarly as above, we could adjust our algorithm to yield drawings with high total resolution by simply taking into account also the angular resolution of the drawing. In particular, if the total resolution of the drawing is defined by its angular resolution, then the way we compute the critical vertices of this drawing has to change; the critical vertices must be the endvertices of the pairs of edges that define the angular resolution. Also, at each iteration of our algorithm we have to ensure that the total resolution does not decrease. We do so by rejecting candidate positions which yield a reduced total~resolution.

@@

In this section, we present the results of our experimental evaluation. For comparison purposes, apart from our algorithm, we also implemented the force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. The implementations\footnote{Our implementation is publicly available at \url{https://github.com/michaelbekos/crossingresolution}} were in Java using yFiles~\cite{DBLP:books/sp/04/WieseE004}. The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM.
%
As a test set for our experiment, we used the non-planar Rome graphs~\cite{DBLP:reference/crc/BattistaD13}, which form a collection of around 8.100 benchmark graphs; in Appendix~\ref{app:north}, we also report on the AT\&T graphs.

@@

% ==================================================================
\myparagraph{Discussion}
%
% ==================================================================
%
%In this paper, we introduced a new heuristic aiming to produce drawings of high crossing resolution, for which we also presented variants that take into account other common aesthetic criteria in Graph Drawing. Our experimental evaluation indicates that the new heuristic is competitive to state of the art force-directed algorithms, even when restricted to a given maximum aspect ratio. As a future direction, we plan to evaluate the performance of variants of our heuristic that compromise between even more aesthetic criteria.
%
While working on this project, we made some useful observations and  obtained some interesting insights. In particular, there is a recent hypothesis (also supported by experiments) that drawings, in which the crossing angles, are large are easy to read and understand. What we observed is that the drawings that are optimized only in terms of the crossing angles might be arbitrarily bad and may have several undesired properties. In particular, in these drawings it was very common to have adjacent edges to run almost in parallel and vertices to be very close to each other (nearly at the same spot). Hence, the angular resolution and the aspect ratio were very often poor. The additional restrictions that we imposed regarding the angular resolution and the aspect ratio helped significantly improving the readability of the drawings, without loosing too much of their quality in terms of the crossing resolution.
